\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[a4paper, total={6in, 8in}]{geometry}
\usepackage{amsmath, amsfonts}
\usepackage{hyperref, graphicx}

\title{CSCI 145 Project}
\date{}

\begin{document}

\maketitle

\section*{Project}

There are three deliverables for your project.

\section{Project Codebase}
Please \textbf{make your code publicly available} (Github repository
or Colab notebook) and submit the link via Gradescope by
\textbf{11:59pm Friday, December 12}.

I want the majority of your focus to be on writing code.
You are welcome to use LLMS or code from the internet but you \textbf{must acknowledge the source} of code you did not write.

I will evaluate your code with the following criteria:
\begin{itemize}
    \item Your code should solve the problem you chose to work on.
    \item Your code should be error-free if I were to run it.
    \item Your code should be well structured with headings describing 
    the purpose of a block of code.
\end{itemize}

\section{Project Presentation}
Please submit public links to your \textbf{two} recorded
project presentations via Gradescope by \textbf{10am Tuesday, December 9}.

An important part of your work is sharing it with others.
You will have 10 minutes to present your work during the final time slot.
\textbf{I expect you to practice your presentation.}
Your deliverables will include \textbf{two} 10 minute recordings of your presentation.
The second recording should have obvious improvements over the first.

I expect your presentations to include the following:
\begin{itemize}
    \item A problem statement describing what you sought to do.
    \item An accessible description of how you solved the problem.
    \item Two or three lovely visuals that you made from your code.
    (Come to me if you're having trouble thinking of ideas for this).
    \item A discussion of what you learned from the project.
\end{itemize}

\paragraph{Note:} If you are working in a group, I expect each member to
cover one bullet in class.
However, when you are submitting the two recordings of the presentation,
I expect each group member to submit their own recordings of the full
presentation.

\section{Project Report}
Please submit your written responses via Gradescope
by \textbf{11:59pm Friday, December 12}.

A short talk is often not enough to convey the work you did.
Please expand on the bullets in the project presentation with
at least three sentences per prompt.

\paragraph{Note:} If you are working a group, I expect each member to submit
a \textbf{different} project report. However, you may share the visualizations.

\end{document}
