\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[a4paper, total={6in, 8in}]{geometry}
\usepackage{amsmath, amsfonts}
\usepackage{hyperref, graphicx}
    

\title{CSCI 145 Project Proposal}
\author{} % Put your name here
\date{}

\begin{document}

\maketitle

\subsection*{Submission Instructions}
Please upload your proposal to Gradescope by \textbf{10am Tuesday, December 2}.

\subsection*{Project Proposal}

Your course project (worth 20\% of your grade) is an opportunity
for you to explore a machine learning topic that is
interesting to you.
You can work on the project by yourself or with a partner.
The purpose of the proposal is to get you thinking 
about your project.

Peruse the topics on the homepage of the course and choose several that are interesting to you.

Start brainstorming how you would go about implementing an algorithm that we discuss.
Think about:
\begin{itemize}
    \item What data sets would you run the algorithm on?
    \item How would you evaluate the algorithm?
    \item What algorithms could you compare your algorithm to?
    \item What interesting visualizations could you make describing your analysis?
    \item How could you improve the algorithm?
\end{itemize}

\subsubsection*{Deliverables}

You should submit the following in a PDF produced from
LaTeX source code:
\begin{itemize}
    \item Your name (and your partner's name if you want to work in a pair)
    \item A tentative title of your project
    \item The topic and algorithm you plan to work on
    \item Answers to the questions above 
\end{itemize}

\end{document}
